---
title: "Data Preparation"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format: 
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```
```{r}
#| echo: false

options(scipen=999)

source(fs::path(here::here(),"2_longitudinal_metabolomics","!libraries.R"))
source(fs::path(here::here(),"2_longitudinal_metabolomics","!directories.R"))

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
#| label: shorten-long-path
#| include: false

create_shortcut <- function(path, short_cut, drive)
{
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path))
}

if(nchar(dir_data) > 80)
{
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))
  
  dir_objects_names <- ls(pattern = "^dir_")
  
  dir_list <- mget(dir_objects_names)
  
  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)  
}

rm(dir_objects_names, short_cut, dir_list, dir_list_shortened)
```

## Data Loading

**Two datasets were loaded in this project. One is plasma metabolomics data, another is Exposure Outcome Data**

```{r load data}
# Load metabolomics data----
met_fts <- read_rds(fs::path(
  dir_metab,
  "plasma_metabolomics_fts.rds"))

# Load exposure outcome data----
tl <- readRDS(fs::path(
  dir_tl_data,
  "tl_covariates_outcomes_environmentalchemicals_l.RDS"))
```

<!-- # ```{r} -->
<!-- # pfas_burden_score <- read.csv( -->
<!-- #   fs::path( -->
<!-- #     dir_data, -->
<!-- #     "Estimated PFAS Burden.csv" -->
<!-- #   ) -->
<!-- # ) -->
<!-- # colnames(pfas_burden_score) <- c("key", "PFAS_burden_deciles", "SE_deciles", -->
<!-- #                                  "PFAS_burden_deciles_summed", "SE_summed_deciles") -->
<!-- # pfas_burden_score$key <- as.character(pfas_burden_score$key) -->
<!-- # ``` -->

## Data Preprocessing

```{r}
covariates <- c("age_baseline", "sex", "race_binary", "parents_income_new", "site")

chemical_names <- c("PFOS", "PFHxS", "PFHpS", "PFOA", "PFNA", "PFDA", "PFHpA", "PFUnDa")

chemical_variables <- c(
  "pfos_untargeted_plasma",
  "pf_hx_s_untargeted_plasma",
  "pf_hp_s_untargeted_plasma",
  "pfoa_untargeted_plasma",
  "pfna_untargeted_plasma",
  "pfda_untargeted_plasma",
  "pf_hp_a_untargeted_plasma",
  "pf_ud_a_untargeted_plasma"
)
```

```{r exposure filling}
tl %>% 
  dplyr::filter(visit == 1) %>%
  dplyr::select(key, visit, all_of(chemical_variables)) %>%
  dplyr::filter(
    rowSums(is.na(across(all_of(chemical_variables)))) < length(chemical_variables)) %>%
  dplyr::filter(
    !grepl("GB|gb", key)
  ) %>%
  drop_na(key, visit) %>% 
  dplyr::select(key) %>% unlist -> subject_w_pfas

tl <- tl %>% 
  dplyr::filter(key %in% subject_w_pfas)
```

```{r}
subject_w_pfas <- tl %>%
  dplyr::select(key, visit) %>% 
  drop_na(key, visit)

subject_w_plasma <- met_fts[[1]] %>% 
  dplyr::select(key, visit) %>% 
  dplyr::filter(
    !grepl("GB|gb", key)
  ) %>% 
  drop_na(key, visit)

subject_include <- inner_join(
  subject_w_pfas, subject_w_plasma,
  by = c("key", "visit")
) %>% 
  dplyr::mutate(visit = factor(visit, levels = c("1", "6", "12", "36"), ordered = T)) %>% 
  arrange(key, visit)
```

I filtered out the rows where all selected PFAS variables (`r paste(chemical_variables, collapse = ", ")`) are missing while keeping at least one non-missing value in the TeenLABS dataset of long format. There were `r dim(subject_w_pfas)[1]` records from `r length(unique(subject_w_pfas$key))` participants remained.

There were `r dim(subject_w_plasma)[1]` records in the plasma metabolomics data, from `r length(unique(subject_w_plasma$key))` participants.

There were **`r dim(subject_include)[1]`** records from **`r length(unique(subject_include$key))`** participants having both untargeted PFAS and plasma metabolomics data.

```{r subset data}
tl_final <- subject_include |>
  left_join(
    tl %>% 
      dplyr::mutate(visit = factor(visit, levels = c("1", "6", "12", "36"), ordered = T)),
    by = c("key", "visit"))
met_fts_final <- modify(
  met_fts,
  ~ subject_include |>
    left_join(
      .x %>% 
        dplyr::mutate(visit = factor(visit, levels = c("1", "6", "12", "36"), ordered = T)),
      by = c("key", "visit"))
)

length(unique(tl_final$key))
```

### Covariates Creation

```{r}
tl_final <- tl_final %>%
  dplyr::mutate(age_baseline = case_when(visit_new == 0 ~  agemos))

tl_final <- tl_final %>%
  dplyr::group_by(key) %>%
  fill(age_baseline) %>% 
  dplyr::ungroup()

tl_final$age_baseline <- tl_final$age_baseline/12

tl_final <- tl_final %>%
  dplyr::group_by(key) %>%
  fill(parents_income) %>% 
  dplyr::ungroup()
tl_final <- tl_final %>% 
  dplyr::mutate(parents_income_new = ifelse(is.na(parents_income), "unknown", parents_income))
tl_final$parents_income_new <- as.factor(tl_final$parents_income_new)
tl_final$parents_income_new <- relevel(tl_final$parents_income_new, ref = "less than 25000")
```

```{r}
sapply(
  tl_final[,c(covariates)],
  typeof
)

tl_final <- tl_final %>% 
  dplyr::mutate(
    sex = factor(sex, labels = c("Male", "Female")),
    race_binary = factor(race_binary, labels = c("Others", "White"))
  )
```

### Mixture Variables

```{r}
# tl_final <- left_join(
#   tl_final,
#   pfas_burden_score[,c("key", "PFAS_burden_deciles", "PFAS_burden_deciles_summed")],
#   by = "key"
# )

tl_final <- tl_final %>% 
  dplyr::mutate(
    pfas_sulfonic_acids = rowSums(across(c(
      pfos_untargeted_plasma, pf_hx_s_untargeted_plasma, pf_hp_s_untargeted_plasma)),
      na.rm = T
    ),
    pfas_carboxylic_acids = rowSums(across(c(
      pfoa_untargeted_plasma, pfna_untargeted_plasma, pfda_untargeted_plasma,
      pf_hp_a_untargeted_plasma, pf_ud_a_untargeted_plasma)),
      na.rm = T
    )
  )
```

## Working Data Creation

```{r data combination}
save(
  tl_final,
  met_fts_final,
  file = fs::path(dir_data, "tl_long_met_untargeted_pfas_workdata.RData")
)
```
