---
title: "MWAS analysis"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format: 
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options: 
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```
```{r}
#| echo: false

options(scipen=999)

source(fs::path(here::here(),"2_longitudinal_metabolomics","!libraries.R"))
source(fs::path(here::here(),"2_longitudinal_metabolomics","!directories.R"))
source(fs::path(here::here(),"2_longitudinal_metabolomics","!functions.R"))

library(nlme)
library(doParallel)
library(foreach)
library(kableExtra)

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
#| label: shorten-long-path
#| include: false

create_shortcut <- function(path, short_cut, drive)
{
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path))
}

if(nchar(dir_data) > 80)
{
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))
  
  dir_objects_names <- ls(pattern = "^dir_")
  
  dir_list <- mget(dir_objects_names)
  
  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)  
}

rm(dir_objects_names, short_cut, dir_list, dir_list_shortened)
```

## Model Parameters

```{r}
chemical_names <- c("PFOS", "PFHxS", "PFHpS", "PFOA", "PFNA", "PFDA", "PFHpA", "PFUnDa")

chemical_variables <- c(
  "pfos_untargeted_plasma",
  "pf_hx_s_untargeted_plasma",
  "pf_hp_s_untargeted_plasma",
  "pfoa_untargeted_plasma",
  "pfna_untargeted_plasma",
  "pfda_untargeted_plasma",
  "pf_hp_a_untargeted_plasma",
  "pf_ud_a_untargeted_plasma"
)

covariates <- c("age_baseline", "sex", "race_binary", "parents_income_new", "site")

mix_variables <- c(
  "PFAS_burden_deciles_summed",
  "pfas_sulfonic_acids",
  "pfas_carboxylic_acids"
)
```

## Data Loading

```{r load data}
load(
  fs::path(
    dir_data,
    "tl_long_met_untargeted_pfas_workdata.RData"
  )
)

platforms <- names(met_fts_final)
```

## LMM-based MWAS

```{r}
#| eval: false

start_time <- Sys.time()

lmm_random_intercept_knot1_covars_PFAS_output <- {}
for(i in 1:length(platforms))
{
  print(i)
  model_df <- met_fts_final[[i]] %>% 
    dplyr::select(-key, -visit)
  covariates_df <- tl_final[,covariates]
  key <- tl_final$key
  visit <- tl_final$visit_year
    
  output_by_chemical <- {}
  for(k in 1:length(chemical_variables))
  {
    print(chemical_variables[k])
    x <- tl_final[,chemical_variables[k]]
    cl <- makeCluster(detectCores()-2)
    registerDoParallel(cl)
    model_output <- foreach(j = 1:dim(model_df)[2], .combine = rbind, .packages = c("nlme")) %dopar%
      fit_lmm_random_intercept_covars_JT(x = x, y = model_df[,j], key = key, visit = visit, covariates_df)
    stopCluster(cl)
    
    model_output <- model_output %>%
      as.data.frame %>%
      mutate_at(vars(coef, se, statistic, pvalue, pvalue_interaction, L_ratio), as.numeric)
    
    model_output$feature <- rep(colnames(model_df), each = 2)
    
    model_output$p_adjust_interaction <- p.adjust(model_output$pvalue_interaction, method = "BH")
    
    output_by_chemical[[k]] <- model_output
    
    rm(model_output, x)
  }
  
  lmm_random_intercept_knot1_covars_PFAS_output[[i]] <- output_by_chemical |>
    bind_rows()
  
  rm(output_by_chemical, model_df)
}
names(lmm_random_intercept_knot1_covars_PFAS_output) <- platforms

save(lmm_random_intercept_knot1_covars_PFAS_output,
     file = fs::path(
       dir_results |> dirname(), 
       "1_code",
       "2_longitudinal_metabolomics",
       "temp_data",
       "model_statistics_untargeted_PFAS_longitudinal_metabo.RData"))

end_time <- Sys.time()
end_time - start_time
```

```{r}
load(
  fs::path(
    dir_results |> dirname(),
    "1_code",
    "2_longitudinal_metabolomics",
    "temp_data",
    "model_statistics_untargeted_PFAS_longitudinal_metabo.RData"
  )
)
```

## Summary

### Raw p-values \< 0.05

```{r}
sig_list <- lapply(
  lmm_random_intercept_knot1_covars_PFAS_output,
  function(df) get_sig_by_chemical(df, "pvalue", 0.05)
) |>
  modify2(
    names(lmm_random_intercept_knot1_covars_PFAS_output),
    ~ .x %>% 
      dplyr::mutate(mode = .y)
  ) |>
  bind_rows() %>% 
  dplyr::mutate(
    exposure = factor(exposure, levels = chemical_variables)
  ) %>% 
  dplyr::arrange(mode, exposure)

sig_list$PFAS <- chemical_names[match(sig_list$exposure, chemical_variables)]
```

```{r}
sig_list %>% 
  dplyr::select(mode, PFAS, pvalue) |>
  kbl(
    col.names = c("Mode", "PFAS chemicals", "# of Sig (p < 0.05)")
  ) |>
  kable_paper("hover")
```

### PCA-corrected p-values

```{r get mode-wide adjusted p-values}
#| eval: false

p_threshold_pca <- modify(
  met_fts_final,
  ~ {
    pca_mwas <- prcomp(.x[,-c(1:2)])
    eigenvalues_mwas <- pca_mwas$sdev^2
    M_eff_mwas <- sum(eigenvalues_mwas > 1)
    0.05/M_eff_mwas
  }
)

# return adjusted p threshold 0.000085
```

```{r}
sig_list <- lapply(
  lmm_random_intercept_knot1_covars_PFAS_output,
  function(df) get_sig_JT_by_chemical(df, "pvalue_interaction", 0.000085)
) |>
  modify2(
    names(lmm_random_intercept_knot1_covars_PFAS_output),
    ~ .x %>% 
      dplyr::mutate(mode = .y)
  ) |>
  bind_rows() %>% 
  dplyr::mutate(
    exposure = factor(exposure, levels = chemical_variables)
  ) %>% 
  dplyr::arrange(mode, exposure)

sig_list$PFAS <- chemical_names[match(sig_list$exposure, chemical_variables)]
```

```{r}
sig_list %>% 
  dplyr::select(mode, PFAS, pvalue_interaction) |>
  kbl(
    col.names = c("Mode", "PFAS chemicals", "# of Sig (nominal p < 0.000085)")
  ) |>
  kable_paper("hover")
```

```{r}
sig_list %>% 
  dplyr::select(mode, PFAS, pvalue_interaction) %>% 
  dplyr::group_by(PFAS) %>% 
  dplyr::reframe(sum = sum(pvalue_interaction, na.rm = T))
```

## Manuscript

I will use nominal p values first with 0.000085 as thresholds.

```{r extract significant features}
sig_feature_list <- lapply(
  lmm_random_intercept_knot1_covars_PFAS_output,
  function(df) get_sig_JT_features_by_chemical(df, "pvalue_interaction", 0.000085)
)

saveRDS(
  sig_feature_list,
  file = fs::path(
    dir_results |> dirname(),
    "1_code",
    "2_longitudinal_metabolomics",
    "temp_data",
    "sig_JT_feature_by_chemical_pca_corrected.rds"
  )
)
```
